---
title: "ACS Exploration"
author: "Todd Suomela"
date: "11/21/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("acs")
```

# Initial Explorations of the ACS package using the Kansas data provided

These are taken from the [package documentation](https://cran.r-project.org/web/packages/acs/acs.pdf) at CRAN.

```{r}
data("kansas09")
class(kansas09)
str(kansas09)
dim(kansas09)
dim(kansas07)
kansas09[1:3, 2:4]
# Slicing the matrix by rows (geographies) 1:3 and columns (variables) 2:4

```

## Some plots generated based on ACS package docs

```{r}
plot(kansas09[4,4])
# Plot for a single variable in a single county.

plot(kansas07[,10])
# Plot for single column (variable) ,10 by all counties in geography (7 counties)

plot(kansas07[4,3:10])
# Plot for a single county, across variables 3:10.

kansas07[4,3:10]
# Print out for previous plot


plot(kansas07[7,10])
# Plot single geography (county), single column (variable), with standard error bars.

plot(kansas07[7,10], col="blue", err.col="purple", err.lty=3)
# Modifying the last plot with some potential flags.


plot(kansas07[7,49], col="lightblue", type="h", x.res=3000,
     err.col="purple", err.lty=3, err.lwd=4, conf.level=.99,
     main=(paste("Distribution of Females>85 Years in ",
                 geography(kansas07)[7,1], sep="")),
     sub="(99-percent margin of error shown in purple)")

```

## Accessor Functions


```{r}
geography(kansas09)
acs.colnames(kansas09)
acs.units(kansas09)
```

The interesting column for the following experiments is 50, Total Population
kansas09[1, Universe...TOTAL.POPULATION.IN.THE.UNITED.STATES..Total]

## Doing some subsetting

```{r}
summary(kansas07[,50])

```


## Working with estimates

Working again with column 50 total population.

```{r}
estimate(kansas09[,50])
dim(estimate(kansas09[,50]))
class(estimate(kansas09[,50]))

# This is the sum method provided by the ACS package, which includes confidence intervals. 
sum(kansas09[,50])

# This just sums the values of the estimates, ignoring the errors.
sum(estimate(kansas09[,50]))

# sum all rows, across the columns for male population [3:25]
sum(kansas09[, 3:25], agg.term=c("Kansas","Total Males"))

# adding optional agg.term
sum(kansas09[,50], agg.term = c("Kansas","Total Population"))

mean(estimate(kansas09[,50]))
as.vector(estimate(kansas09[,50]))
as.vector(estimate(kansas07[,50]))
boxplot(as.vector(estimate(kansas07[,50])))

```

