---
title: "Untitled"
author: "Todd Suomela"
date: "10/5/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions for supporting the primary analysis

The goal of this document is to develop a set of functions that will allow someone to merge all of the demographic and primary vote data into a form which will allow for basic choropleth and linear model testing.

Before using the functions you must make sure that the df_county_demographics, county.regions, and state primary file are both loaded into R.

## Step 0: Loading data from file

User needs to define the state name.

```{r}

state.for.analysis <- "pennsylvania"

# Utility function for converting strings to Title Case
upper_name <- function(x) {
  s <- strsplit(x, " ")[[1]]
  paste(toupper(substring(s, 1,1)), substring(s, 2),
      sep="", collapse=" ")}

read_data <- function(x) {
  uname <- upper_name(x)
  data_file <- paste("data/", uname, " Primary Results by County.csv", collapse = "", sep = "")
  incoming_data <- read.csv(data_file, header = TRUE)
  subset(incoming_data, incoming_data[1] != "") # subset to drop data where first column is empty
}

state.vote.data <- read_data(state.for.analysis)

# conversions to lower case for standardized processing
names(state.vote.data) <- tolower(names(state.vote.data))
state.vote.data$county <- tolower(state.vote.data$county)

```
## Step 1: Filter and standardize choroplethr data for joining.

First load necessary libraries for county data frames and choropleth functions, as well as dplyr for data cleaning.

Filter the county.regions df from the choroplethr package for the state of interest.

You need to filter the county.regions df first because of potential duplicate county names across the nation.

```{r}
library(choroplethr)
library(choroplethrMaps)
library(dplyr)

data("county.regions")
data("df_county_demographics")

counties.geo <- filter(county.regions, state.for.analysis == county.regions$state.name)

```

## Step 2: Tidying the data

```{r}
# make sure the first column of the state.vote.data has a standard name that can be merged with counties.geo
names(state.vote.data)[1] <- "county.name"

tidy.primary <-
  state.vote.data %>%
  tidyr::gather(candidate, votes, -1) # using -1 to ignore first column of data frame which contains the county or subdivision name


```



## Step 3: Perform your joins

This is merging data from the state.vote.data and counties.geo data frames so that choroplethr function can be performed. The join is performed on matches for the county.name column.

```{r}
# joining counties.geo and state.vote.data
full.geo.demo.data <-
  left_join(tidy.primary, counties.geo, by = "county.name") %>%
  left_join(df_county_demographics, by = "region")
```


## Step 4: Create proportions for demographic data and votes

Note that percent2proportion must be called by quoting the column names. For example, "percent_asian" instead of percent_asian.


```{r}
full.geo.demo.data <-
  mutate(full.geo.demo.data, vote.proportion = votes / total_population)

# finally a one line hack that transforms all of the percent race items into proportions. Not ideal but it will do the job.

full.geo.demo.data[c("proportion.white",
                     "proportion.black",
                     "proportion.asian",
                     "proportion.hispanic")] <-
                     sapply (full.geo.demo.data[c("percent_white",
                     "percent_black",
                     "percent_asian",
                     "percent_hispanic")],
                     function(x)
                     x / 100)

# I would prefer to generalize this as a function but so far I have not been successful.
# Reference for this function [stackoverflow](http://stackoverflow.com/questions/26003574/r-dplyr-mutate-use-dynamic-variable-names#31689987)
# This function takes a data frame and the name of the variable column you wish to convert from percent to proportion. It also adds a new column with the name proportion_%race% instead of percent_%race%

# percent_to_proportion <- function(df, variable) {
#     varname <- gsub("percent", "proportion", variable)
#     for (i in length(variable)) {
#       assign(varname, variable / 100)
#       # mutate (df, varname = variable[i] / 100)
#       # df[[varname[i]]] <- df[[variable[i]]] / 100
#     }
#     df
# }
# 
# # An early attempt which worked as long as I passed the variable and data frame in four different calls for each percent_race
# percent_to_proportion <- function (df, per) {
#   df[[per]] / 100
# }

```

## Step 5: Making maps for candidates in the current data frame

Call choroplethr for each candidate

```{r}

dplyr::rename(full.geo.demo.data, value = vote.proportion) %>%
  group_by(candidate)
# %>%
#   mean(value)


```

# Step 6: exploratory data analysis

```{r}
library(ggplot2)

hist_vote.proportion <-
  ggplot(full.geo.demo.data, aes(vote.proportion))
  + geom_histogram() + facet_wrap( ~ candidate)

sp_votes.population <-
  ggplot(full.geo.demo.data, aes(x = votes, y = total_population, color = candidate)) + geom_point()

sp_votes.population_smallpop <-
  filter(full.geo.demo.data, total_population < 50000) %>%
  ggplot(. , aes(x = votes, y = total_population, color = candidate)) + geom_point()

```

The histogram shows the number of counties that each candidate won binned by vote proportion.  So Jeb Bush and Marco Rubio had very small vote proportions across almost all counties.

For the state of Pennsylvania the county with the largest population cast 200,000+ votes for Clinton; 100,000+ for Sanders; and ~ 25,000 for Trump.  As the population of the county decreases, Trump gets larger numbers of votes.

The second scatter plot covers counties with populations less than 50000 and shows a clear Trump advantage in counties with smaller populations.

